Monster or Lambda?  A likelihood analysis

### 2023-01-24 inst/home_advantage.Rmd
show complex form of same dataset for reference
fixes #196

Here I analyse a small synthetic dataset for home ground advantage,
then go on to discuss a dataset presented by Agresti.  First generate
the dataset by hand:

@@

away_games_won <- matrix(c(
    NA, 05, 02, 02,
    02, NA, 10, 02,
    00, 04, NA, 07,
    02, 06, 04, NA),
    nrow=4,ncol=4,byrow=TRUE)

@@

Thus `home_games_won[1,2] == 16` means A played at home against B and
won 16 times; `home_games_won[2,1] == 19` means B played at home
against A and won 19 times.  Also, `away_games_won[1,2] == 5` means A
played away against B and won 5 times, and `away_games_won[2,1] == 2`
means B played away against A and won 2 times.  Alternatively, A
played B 16+19+5+2=45 times; A played at home 16+2=18 times and won 16
times and lost 2 times; and B played at home 19+5=24 times and won 19
times and lost 5 times.

@@

```{r,label=makehyper2,cache=TRUE}
H <- home_away(home_games_won,away_games_won)
H
options("use_alabama" = TRUE) # to avoid the stupid wmmin bug
specificp.gt.test(H,"home",0)
```

@@

```{r,label=showvisH}
dotchart(maxp(H))
```

@@

```{r,label=showprofilelikelihood, cache=TRUE}
home_strength <- seq(from=0.2,to=0.6,length=20)
plot(home_strength,profsupp(H,"home",home_strength),type="b",pch=16,
     xlab="home strength",ylab="support",main="profile likelihood for home advantage")
abline(h=c(0,-2))
```

@@

```{r,label=baseballtransplot,cache=TRUE}
ordertransplot(baseball_maxp[1:7],maxp(baseball)[1:7],xlab="old",ylab="new",plotlims=c(0.1,0.16))
```

>>>>

Here I analyse a small synthetic dataset for home ground advantage,
then go on to discuss a dataset presented by Agresti.  But first, the
smallest nontrivial example is a 2x2 case:

```{r,label=smallestexample}
M <- matrix(c(
    NA, 1+5i, 3+11i, NA),
    nrow=2,ncol=2,byrow=TRUE)

teams <- LETTERS[seq_len(2)]
dimnames(M) <- list("@home" = teams,"@away"=teams)
dimnames(M) <- list("@home" = teams,"@away"=teams)
M
home_away(M)
```

Above we see (from `M[1,2]`) that A played at home $1+5=6$ times,
winning 1 and losing 5; and B played at home $3+11=14$ times, winning
3 and losing 11.   A slightly more complicated example:

@@

away_games_won <- matrix(c(
    NA, 05, 02, 02,
     9, NA, 10, 02,
     3, 04, NA, 07,
     8, 06, 04, NA),
    nrow=4,ncol=4,byrow=TRUE)

@@

Thus `home_games_won[1,2] == 16` means A played at home against B and
won 16 times; `home_games_won[2,1] == 19` means B played at home
against A and won 19 times.  Also, `away_games_won[1,2] == 5` means A
played away against B and won 5 times, and `away_games_won[2,1] == 2`
means B played away against A and won 2 times.  Alternatively, A
played B 16+19+5+2=45 times; A played at home 16+2=18 times and won 16
times and lost 2 times; and B played at home 19+5=24 times and won 19
times and lost 5 times.  Alternatively we may use complex numbers to
represent the same dataset:

```{r,label=showhomeawayusingcomplex}
home_games_won + 1i*away_games_won
```

@@

```{r,label=makehyper2,cache=TRUE}
H <- home_away(home_games_won,away_games_won)
H
options("use_alabama" = FALSE) # to avoid the stupid wmmin bug
specificp.gt.test(H,"home",0)
options("use_alabama" = TRUE) # to avoid the stupid wmmin bug
```

@@

```{r,label=showvisH,cache=TRUE}
(mt <- maxp(H))
```

@@

```{r,label=showprofilelikelihood, cache=TRUE}
home_strength <- seq(from=0.15,to=0.45,length=13)
plot(home_strength,profsupp(H,"home",home_strength),type="b",pch=16,
     xlab="home strength",ylab="support",main="profile likelihood for home advantage")
abline(h=c(0,-2))
```

@@

```{r,label=baseballtransplot,cache=TRUE}
ordertransplot(baseball_maxp[1:7],maxp(baseball)[1:7],xlab="old",ylab="new",plotlims=c(0.01,0.16))
```

### 2023-01-24 inst/monster_vs_lambda.Rmd
fixes #196

```{r}
`ha3` <- function(home_games_won,away_games_won,lambda){

@@

    H <- hyper3(pnames=c(teams,"M","lambda"))

    for(i in seq_len(nrow(home_games_won))){
        for(j in seq_len(ncol(home_games_won))){
            if(i != j){  # diagonal means a team plays itself, meaningless.
                ## First deal with home_games:
                game_winner <- teams[i]
                game_loser  <- teams[j]
                no_of_matches <- home_games_won[i,j] # won with home strength helping

                jj <- c(lambda,1)
                names(jj) <- c(game_winner,"M")
                H[jj] %<>% inc(no_of_matches)  # won with home strength helping

                jj <- c(lambda,1,1)
                names(jj) <- c(game_winner,game_loser,"M")
                H[jj] %<>% dec(no_of_matches)

                ## now away games:
                no_of_matches <- away_games_won[i,j] # won without the benefit of home strength
                jj <- 1
                names(jj) <- game_winner
                H[jj] %<>% inc(no_of_matches)  # won without home ground helping...

                jj <- c(1,lambda,1)
                names(jj) <- c(game_winner,game_loser,"M")
                H[jj] %<>% dec(no_of_matches)  # home strength nevertheless on denominator

            } # if(i != j) closes
        } # j loop closes
    } # i loop closes
    return(H)
}
```

(maybe one day I will include this in the `hyper2` package).  Copying
from `home_advantage.Rmd`:

```{r label=definehomeawaymatrix}
D  <- matrix(c(
    NA, 16, 12, 11, 19, NA, 19, 16,
    17, 12, NA, 11, 11, 12, 12, NA),
    nrow=4,ncol=4,byrow=TRUE) + 1i* matrix(c(
    NA, 05, 02, 02, 02, NA, 10, 02,
    00, 04, NA, 07, 02, 06, 04, NA),
    nrow=4,ncol=4,byrow=TRUE)

teams <- LETTERS[1:4]
dimnames(D) <- list("@home" = teams,"@away"=teams)
D
```

```{r,tryha3,cache=TRUE}
H <- ha3(D,lambda=1.88)
s <- specificp.gt.test(H,"M",0,n=1)  # NB lambda = 1.88
s
```

@@

Above but in one line:

```{r,label=trytwovalues,cache=TRUE}
ll <- function(lambda){specificp.gt.test(ha3(D,lambda=lambda),"M",0,n=1)$null_support}
ll(1.22)
ll(1.88)
```

```{r,label=calclam,cache=TRUE}
lam <- seq(from=2.5,to=6,len=15)
supp <- sapply(lam,ll)
```

```{r}
plot(lam,supp-max(supp),type='b')
```

>>>>

```{r defineha3}
`ha3` <- function(home_games_won,away_games_won,lambda){  # based on home_away3()

@@

    H <- hyper3(pnames=c(teams,"lambda","M"))

    for(i in seq_len(nrow(home_games_won))){
        for(j in seq_len(ncol(home_games_won))){
            if(i != j){  
                home_team <- teams[i]
                away_team <- teams[j]

		home_wins <- home_games_won[i,j]
		away_wins <- away_games_won[i,j] 
                no_of_matches <- home_wins + away_wins

                ## home wins:
                jj <- c(lambda,1)
                names(jj) <- c(home_team,"M")
                H[jj] %<>% inc(home_wins)

                ## away wins:
                jj <- 1
                names(jj) <- away_team
                H[jj] %<>% inc(away_wins)

                ## denominator
                jj <- c(lambda,1,1)
                names(jj) <- c(home_team,away_team,"M")
                H[jj] %<>% dec(no_of_matches)
            } # if(i != j) closes
        } # j loop closes
    } # i loop closes
    return(H)
}
```

We will try on a simple case:

```{r,label=try2x2}
M <- matrix(c(
    NA, 1+5i, 3+11i, NA),
    nrow=2,ncol=2,byrow=TRUE)

teams <- LETTERS[seq_len(2)]
dimnames(M) <- list("@home" = teams,"@away"=teams)
dimnames(M) <- list("@home" = teams,"@away"=teams)
M
home_away(M)
ha3(M,lambda=1.88)
```

(maybe one day I will include this function in the `hyper2` package).
Copying from `home_advantage.Rmd`:

```{r label=definehomeawaymatrix}
home_games_won <- matrix(c(
    NA, 16, 12, 11,
    19, NA, 19, 16,
    17, 12, NA, 11,
    11, 12, 12, NA),
    nrow=4,ncol=4,byrow=TRUE)

away_games_won <- matrix(c(
    NA, 05, 02, 02,
     9, NA, 10, 02,
     3, 04, NA, 07,
     8, 06, 04, NA),
    nrow=4,ncol=4,byrow=TRUE)

teams <- LETTERS[1:4]
dimnames(home_games_won) <- list("@home" = teams,"@away"=teams)
dimnames(away_games_won) <- list("@home" = teams,"@away"=teams)

D <- home_games_won + 1i*away_games_won
D
```

Above we see, from row 1 column 2 [`36+10i`], that A played at home
against B a total of $36+10=46$ times, winning $36$ and losing $10$;
From row 2 column 1 [`11+10i`] we see that B played at home against A
a total of 21 times, winning 11 and losing 10.  We can use the Monster
likelihood function and test $H_0\colon M=0$:

```{r,tryha3,cache=TRUE}
H <- ha3(D,lambda=1.01)
print(H)
s <- specificp.gt.test(H,"M",0,n=10)  # NB lambda = 1.88
s
```

We see strong evidence that $M>0$.

@@

We want to plot a profile log-likelihood function, that is, a
likelihood function corresponding to the evaluate constrained to
having a particular value of $M$:

```{r,label=marginallambdaprofilelike,cache=TRUE}
f <- function(lambda){
 write(date(),file="marginallambdaprofilelike",append=TRUE)  # so we can check its progress
 return(maxp(ha3(D,lambda=lambda),give=TRUE)$value)
}

lam <- seq(from=1.2,to=4.6,len=15)
supp_M_eq_0 <- sapply(lam,f)
```

```{r,plotprofilelikem0}
plot(lam,supp_M_eq_0,type="b",xlab="lambda")
```

Now enforce the constraint $\lambda=1$ [and allow $M$ to vary],
thereby reducing the likelihood function from a `hyper3` to a `hyper2`
object:

```{r,tryiconprofsupp,cache=TRUE}
(H <- home_away(D))  # observe that the monster is 'home'
home <- seq(from=0.15,to=0.4,len=15)
supp_lambda_eq_0 <- profsupp(H,"home",home,relative=FALSE)
plot(home,supp_lambda_eq_0 - max(supp_lambda_eq_0),type='b')
abline(h=c(0,-2))
```

```{r,showmaxplam1,cache=TRUE}
maxp(H)
```

```{r,label=plotordinarysupport}
plot(lam,supp_M_eq_0-max(supp_M_eq_0),type='b',main='Profile likelihood with M=0')
grid()
plot(home,supp_lambda_eq_0 - max(supp_lambda_eq_0),type='b')
```

```{r,label=plotbothsupports}
both <- c(supp_lambda_eq_0,NA,supp_M_eq_0)
plot(both-max(both,na.rm=TRUE),type='b',col=c(rep("black",length(supp_lambda_eq_0)),"green",rep("red",length(supp_M_eq_0))))
abline(h=c(0,-2))
```

Support difference:

```{r,label=supportdifference}
max(supp_M_eq_0)-max(supp_lambda_eq_0)
```

exceeding Edwards's two-units-of-support criterion by a comfortable
margin.  Note that both have equal numbers of degrees of freedom.

# Create a new dataset by direct simulation

```{r,syntheticdataP}
set.seed(0)
P <- matrix(0,4,4)
teams <- LETTERS[1:4]
dimnames(P) <- list("@home" = teams,"@away"=teams)
P
n <- 2000 # number of matches between a pair 
M <- 0.0005  # home monster
strengths <- 1/(1:4) # Zipf
strengths <- strengths/sum(strengths)
names(strengths) <- teams
strengths
for(home_team in 1:4){
  for(away_team in 1:4){
    if(home_team != away_team){
      prob_home_win <- (strengths[home_team] + M)/(strengths[home_team] + strengths[away_team] + M)
      home_wins <- rbinom(1,n,prob_home_win)
      home_losses <- n - home_wins
      P[home_team,away_team] <- home_wins + 1i*home_losses
    }
  } 
}
diag(P) <- NA
P
HP <- home_away(P)
```

```{r,label=showlike}
HP
```

```{r,label=maxpmonsteqq,cache=TRUE}
mH <- maxp(HP,n=3,give=T)
mH
```
